---
title: "Assignment 6: GLMs (Linear Regressions, ANOVA, & t-tests)"
author: "Emma DeAngeli"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()

library(tidyverse)
library(agricolae)
library(ggplot2)
library(lubridate)
# other needed packages

ChemPhys <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

ChemPhys$sampledate <- as.Date(ChemPhys$sampledate, format = "%m/%d/%Y")


#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "darkblue"), 
        legend.position = "right")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July is the same with depth across all lakes
Ha: Mean lake temperature recorded during July is not the same with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ChemPhys.filtered <- ChemPhys %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()
  

#5
ggplot(ChemPhys.filtered, aes(x = depth, y = temperature_C)) + 
  geom_point() +
  ylim(0, 35) +
  geom_smooth(method = 'lm', se = FALSE, color = "darkgreen")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as depth increases, temperature decreases. The distribution of the points suggests that there many be a linear trend, particularly around depths 2 through 8. However, temperature does not seem to change with depth for the first several meters and then after a certain depth, which could suggest there may be an exponential trend that is more suited to the data.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp.regression <- lm(ChemPhys.filtered$temperature_C ~ ChemPhys.filtered$depth)
summary(temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: There is a significant negative correlation between temperature and depth and the model explains around 73.9% of the total variance in temperature. The residual error is 3.835 on 9726 degrees of freedom. The p-value is less than 2e-1, so it is statistically significant. Per every 1m change in depth, temperature is predicted to decrease by about 1.95 degrees C.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
ChemPhysAIC <- lm(data = ChemPhys.filtered, temperature_C ~ year4 + daynum + 
              depth)
step(ChemPhysAIC)
summary(ChemPhysAIC)


#10
ChemPhysMR <- lm(data = ChemPhys.filtered, temperature_C ~ year4 + daynum + 
              depth)
summary(ChemPhysMR)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using year4, daynum, and depth as explanatory variables to predict temperature. This model explains 74.12% of the observed variance which is an improvement from the previous model's 73.9%.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
ChemPhys.anova <- aov(data = ChemPhys.filtered, temperature_C ~ lakename)
summary(ChemPhys.anova)

ChemPhys.anova2 <- lm(data = ChemPhys.filtered, temperature_C ~ lakename)
summary(ChemPhys.anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Difference in mean temperature is significantly different among lakes for the month of July with a p-value of 1.45e-06. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(ChemPhys.filtered, aes(x = depth, y = temperature_C, color = lakename)) + 
  geom_point(aes(alpha = 0.5)) +
  ylim(0, 35) +
  geom_smooth(method = 'lm', se = FALSE, aes(color = lakename))

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(ChemPhys.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: There was no significant difference in mean temperature between Peter and Paul as well as Peter and Ward. No lake has a mean temperature that is statistically distinct from all other lakes.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could determine pairwise differences and plot the results.


